Measuring bias in (strategically) missing EPA pollution data

### 2021-11-30 paper/5min-presentation/5min-presentation.Rmd
Updating 5 min presentation with map of EPA and PA sensors.

# Motivation
## Clean Air
- The Clean Air Act (1970) established National Ambient Air Quality Standards
(NAAQS) for US counties.

@@

@@

<!-- add space here -->
- NAAQS only requires each monitor to report 75\% of their readings, per quarter.

@@

- How biased is *missing* air pollution data from self-reporting US EPA monitors?

@@

# Project Overview
<!-- :
     latex_engine: xelatex
header-includes:
   - \usetheme[style=dark]{Nord}
   - \usepackage{YGbeamer}
(i) an interesting, well-posed question; 
(ii) an adequate model, such that the question makes sense within the context of
the model; 
(iii) an identification strategy, such that (again within the context of the 
model) the question can be answered; 
“What is your identification strategy?” = “What research design (and assumptions) are you using?”
(iv) a set of proposed hypothesis tests or statistics which map your
data into falsifiable statements that can provide answers to your question. You
should /not/ provide any results; this presentation is about empirical strategy,
not about data-dependent outcomes.
-->

@@

## This project
- Focus on EPA pollution data that is missing *in time*; limited to California.
- Using new consumer-based pollution monitors to understand the bias in EPA data.
- Avoids using satellite estimates (has been shown to have significant error).
<!-- - Understanding the distribution of bias in missing California air pollution data using consumer-based pollution -->
<!-- monitors that have low error near their location. -->

# Purple Air Monitors
[insert maps of California EPA and PA monitors, timelaps GIF? Timeline of adoption]

[insert pictures of PA outdoor monitors]

@@

1. Predict EPA pollution at missing times

2. Estimate bias between predicted pollution at missing times and reported pollution at nonmissing times.

3. Estimate California counties' counterfactual attainment status using included predicted missing pollution data.

@@

\[
EPA_{i,t} = \gamma_{i,0} + \sum\limits_{j\in J_i}\sum\limits_{k=1}^7\gamma_{j,k} PA_{j,t} + u_{i,t}
\]

- Analysis done at the month and quarter level; suppressing that subscript.
- $t$ is a unique hour within a given month or quarter.
- EPA monitor $i$ at time $t$ reads PM2.5 pollution $EPA_{i,t}$.
- For each EPA monitor $i$, there are $J_i$ Purple Air monitors within a 10-mile radius.
[insert diagram of two EPA monitors with PA monitors surrounding them]
- Purple Air monitor $j\in J_i$ at time $t$ reads PM2.5 pollution $PA_{j,t}$.
[insert diagram of one EPA monitor and surrounding PA monitors, with wind directions]

# Models: Hour-by-Day-of-week Bias of Missing EPA Monitor Pollution Data
\def\mm{\mathcal{M}}
\def\nn{\mathcal{N}}
\vspace{-1em}
\begin{align*}
\intertext{\textbf{Missingness Bias:}}
Bias_{i,h,d} &= \frac{1}{|\nn_{i,h,d}|}\sum\limits_{t\in \nn_{i,h,d}}EPA_{i,t} 
- \frac{1}{|\mm_{h,d}|}\sum\limits_{t\in \mm_{h,d}}\widehat{EPA}_{i,t}\\
\text{where } \mm_{i,h,d}&= \{t: t\text{ is at hour $h$ and day $d$ and $EPA_{i,t}$ is Missing}\};\\
  \nn_{i,h,d}&= \{t: t\text{ is at hour $h$ and day $d$ and $EPA_{i,t}$ is Non-missing}\} \\
\uncover<2->{\intertext{We can also define the \textbf{algorithm bias}     as the Hour-by-Day-of-week prediction error}
\widetilde{Bias}_{i,h,d} &= \frac{1}{|\nn_{i,h,d}|}\sum\limits_{t\in \nn_{i,h,d}}EPA_{i,t} 
- \frac{1}{|\mm_{h,d}|}\sum\limits_{t\in \mm_{h,d}}\widehat{EPA}_{i,t} \\}
\uncover<3->{\intertext{We can also define $Bias_{j,h,d}$ and $\widetilde{Bias}_{j,h,d}$ for PA monitor $j$ (we'll come back to this).}}
\end{align*}

@@

\begin{align*}
Attain^{annual}_c &= 1 \text{ if \textbf{reported} annual average PM2.5 below threshold}^*\\
  &= 1[\text{equation here}]\\[1em]
Attain^{daily}_c &= 1 \text{ if 98$^{th}$ percentile of \textbf{reported} daily average PM2.5 below threshold}^*\\
  &= 1[\text{equation here}]\\[1em]
\widehat{Attain}^{annual}_c &= 1 \text{ if \textbf{predicted} annual average PM2.5 below threshold}^*\\[1em]
\widehat{Attain}^{daily}_c &= 1 \text{ if 98$^{th}$ percentile of \textbf{predicted} daily average PM2.5 below threshold}^*
\end{align*}
$^*$averaged over 3 years in NAAQS standard.
[fill in equations and thresholds]

@@

- Assumption: nearby PurpleAir monitors that are good predictors for EPA monitors
during non-missing times will also be good predictors during missing times.
  - Specifically, reasons for EPA data missingness are not correlated with missingness
  or measurement error in PurpleAir data

@@

  - Calculate sum of squared missingness bias and sum of squared algorithm bias for EPA sensor $i$ and PA sensors $j\in J_i$.
  \vspace{0.5em}
  - Ratio$_k$ = sum of squared missingness bias / sum of squared algorithm bias
  \vspace{0.5em}
  - p-value = $\frac{\text{\# of PA sensors in $i$'s radius with Ratio$_j$ larger than Ratio$_i$}}{\text{\# of PA sensors in $i$'s radius}}$

@@

![PurpleAir correction equation for EPA monitor PM2.5 (RH = relative humidity, also measured by PA monitor)](PA_correction_equation.png){width=50%}

@@

>>>>

```{r packages, include=F}
library(tidyverse)
library(docstring)
library(ggplot2)
# library(mapdata)
# Requires rgdal package to be installed, requires gdal to be installed on computer
# Ubuntu 20.04: sudo aptitude install gdal-bin proj-bin libgdal-dev libproj-dev
```

# Motivation
## Clean Air

- The Clean Air Act (1970) established National Ambient Air Quality Standards
(NAAQS) for US counties.

@@

- NAAQS only requires each monitor to report 75\% of their readings, per quarter.

@@

@@

- How biased are EPA monitor-based measures of local air quality?
<!-- - How biased are self-reported EPA pollution estimates? -->
<!-- - How biased is *missing* air pollution data from self-reporting US EPA monitors? -->

@@

# Context
<!-- :
     latex_engine: xelatex
header-includes:
   - \usetheme[style=dark]{Nord}
   - \usepackage{YGbeamer}
(i) an interesting, well-posed question; 
(ii) an adequate model, such that the question makes sense within the context of
the model; 
(iii) an identification strategy, such that (again within the context of the 
model) the question can be answered; 
“What is your identification strategy?” = “What research design (and assumptions) are you using?”
(iv) a set of proposed hypothesis tests or statistics which map your
data into falsifiable statements that can provide answers to your question. You
should /not/ provide any results; this presentation is about empirical strategy,
not about data-dependent outcomes.
-->

@@

## This project
- Using new consumer-based pollution monitors (PurpleAir).

- Focus on EPA pollution data that is missing *in time*.

- Avoids using satellite estimates (has been shown to have significant error).
<!-- - Understanding the distribution of bias in missing California air pollution data using consumer-based pollution -->
<!-- monitors that have low error near their location. -->

<!-- [why California? Maybe should inlcude all EPA monitor - months that have at least one PA monitor within 10 miles] -->

<!-- [insert maps of California EPA and PA monitors, timelaps GIF? Timeline of adoption] -->

<!-- [insert pictures of PA outdoor monitors] -->

```{r geography intersection functions, include=F}
library(sf)
library(spData)

lonlat_to_state <- function(df, lon_col, lat_col,
                            states = spData::us_states,
                            name_col = "NAME") {
  #' State names for lon, lat vectors
  #' 
  #' @description Return a vector of state names for the lon-lat vectors
  #' 
  #' @param df: a data.frame containing lon lat pairs
  #' 
  #' @param lon_col: A character of the column name containing longitude
  #' 
  #' @param lat_col: A character of the column name containing latitude
  #'
  #' @param states:   An sf MULTIPOLYGON object with 50 states plus DC.
  #'
  #' @param name_col: Name of a column in `states` that supplies the states'
  #'           names.
  #' https://stackoverflow.com/a/8751965/16660792
  ## Convert points data.frame to an sf POINTS object
  pointsDF = data.frame(x = df[, lon_col], y = df[, lat_col])
  pts <- st_as_sf(pointsDF, coords = 1:2, crs = 4326)

  ## Transform spatial data to some planar coordinate system
  ## (e.g. Web Mercator) as required for geometric operations
  states <- st_transform(states, crs = 3857)
  pts <- st_transform(pts, crs = 3857)

  ## Find names of state (if any) intersected by each point
  state_names <- states[[name_col]]
  ii <- as.integer(st_intersects(pts, states))
  state_names[ii]
}

lonlat_country_filter <- function(df, lon_col, lat_col,
                                  country='United States',
                                  name_col='name_long') {
  #' Filter for lat-lons in a specific country
  #' 
  #' @description Return the dataframe filtered to inlcude only US points
  #' 
  #' @param df: a data.frame containing lon lat pairs
  #' 
  #' @param lon_col: A character of the column name containing longitude
  #' 
  #' @param lat_col: A character of the column name containing latitude
  #'
  #' @param country: A character of the country, matching name_col from spData::world
  #'
  #' @param name_col: A character of the column name containing the country name,
  #' from spData::world
  #'
  ## Restrict to county boundaries
  country = spData::world %>%
    filter(!!as.symbol(name_col) == country)
  ## Convert points data.frame to an sf POINTS object
  pointsDF = data.frame(x = df[, lon_col], y = df[, lat_col])
  pts <- st_as_sf(pointsDF, coords = 1:2, crs = 4326)

  ## Transform spatial data to some planar coordinate system
  ## (e.g. Web Mercator) as required for geometric operations
  country <- st_transform(country, crs = 3857)
  pts <- st_transform(pts, crs = 3857)

  ## Find names of state (if any) intersected by each point
  ii <- as.integer(st_intersects(pts, country))
  df %>%
    mutate(include = ii) %>%
    filter(include == 1)
}

```

# PM2.5 Air Pollution Monitors

```{r get EPA sensor locations 2016, include=F, cache=T}
sensor_types = c(88101, 88501, 88502)
years = c()

load("/media/a/E/Programming/github/are219/data/epa/epa_monitors/epa_monitors/data_from_api/88101/api_request_daily_88101_2016.Rdata")
df = yr_dt %>% distinct(latitude, longitude, .keep_all = TRUE) %>% mutate(type=88101, NAAQS_used='yes')
load("/media/a/E/Programming/github/are219/data/epa/epa_monitors/epa_monitors/data_from_api/88501/api_request_daily_88501_2016.Rdata")
df2 = yr_dt %>% distinct(latitude, longitude, .keep_all = TRUE) %>% mutate(type=88501, NAAQS_used='no')
load("/media/a/E/Programming/github/are219/data/epa/epa_monitors/epa_monitors/data_from_api/88502/api_request_daily_88502_2016.Rdata")
df3 = yr_dt %>% distinct(latitude, longitude, .keep_all = TRUE) %>% mutate(type=88502, NAAQS_used='no')
epa_data = bind_rows(df, df2, df3) %>%
  filter(type == 88101) %>% 
  distinct(latitude, longitude, .keep_all = TRUE) %>%
  select(longitude, latitude, everything())

```

```{r get PA sensor locations all, include=F}
result = jsonlite::fromJSON("/media/a/E/Programming/github/are219/data/test_data/manual_webpage_download_USmap_2021-11-29.txt")
library(zoo)
# Get start and last_seen dates from all sensors near the US
pa_data2 = as.data.frame(result$data) %>%
  purrr::set_names(result$fields) %>%
  mutate(date_start = lubridate::as_datetime(date_created),
         date_last = lubridate::as_datetime(last_seen),
         month_start = as.yearmon(date_start),
         month_last = as.yearmon(date_last)) %>%
  select(longitude, latitude, everything()) %>%
  arrange(date_start) %>%
  lonlat_country_filter(lon_col='longitude', lat_col='latitude') %>%
  filter(private == 0,  # public sensor
         location_type == 0,  # outdoor sensor
         channel_state != 0)  # has at least one PM sensor
```

```{r Plot the locations, eval=F, include=F}
min_lat = min(pa_data$lat)
max_lat = max(pa_data$lat)
min_lon = min(pa_data$lon)
max_lon = max(pa_data$lon)
print(paste(max_lon, max_lat, min_lon, min_lat))
usa <- map_data('usa', projection='conic', parameters=c(min_lat))

b = 204/255
plot_usmap(fill='white', color='grey') +
  geom_point(data = usmap::usmap_transform(pa_data),   # Purple Air
             aes(x = lon.1, y = lat.1),
             color = rgb(0, b, b), alpha = 1, size=0.4) +
  geom_point(data = usmap::usmap_transform(epa_data),   # EPA
             aes(x = longitude.1, y = latitude.1),
             color = "red", alpha = 0.25, size=2) +
  labs(title = "US EPA & PurpleAir Pollution Monitors",
       subtitle = "Source: EPA, PurpleAir.comJan 1 to Jun 30 2016") +
  theme(legend.position = "right")

```

```{r Monitor Map with AK and HI, include=F}
# devtools::install_github("hrbrmstr/albersusa")
# sudo apt-get install libudunits2-dev
# install.packages('units')
# install.packages('sf')

# Composite map with AK & HI
usa_map = albersusa::usa_composite(proj="laea") %>%
  fortify(.)  # make it usable in ggplot2
epa_trans <- usmap::usmap_transform(epa_data)
pa_trans <- usmap::usmap_transform(pa_data2)

# Plot PA and EPA sensor locations on US map
ggplot() +
  geom_map(data=usa_map, map=usa_map,
                    aes(map_id=id),
                    color="#2b2b2b", size=0.1, fill=NA) +
  geom_point(data = pa_trans, 
             aes(x = longitude.1, y = latitude.1),  # Purple Air
             color = rgb(0, 204/255, 204/255), alpha = 1, size=0.4) +
  geom_point(data = epa_trans, aes(x = longitude.1, y = latitude.1),  # EPA
             color = "red", alpha = 0.25, size=2) +
  labs(title = "US EPA & PurpleAir Pollution Monitors",
       subtitle = "Source: EPA 2016, PurpleAir.com 2015-2021") +
  ggthemes::theme_map()

ggsave('monitor_map_2015-2021.png', width = 6, height = 4)

```

\vspace{-2em}

```{r Map and images, echo=F, warning=F, message=F, fig.width=12, fig.cap='test'}
p_pa = "../../docs/purpleair/PA_outdoor_monitor_border.png"
p_epa = "../../docs/epa/air-monitoring-site_border.jpg"
p_map = 'monitor_map_2015-2021.png'
library(png)
library(jpeg)
library(grid)
library(gridExtra)
img_map = rasterGrob(as.raster(readPNG(p_map)), interpolate = FALSE)
img1 <-  rasterGrob(as.raster(readPNG(p_pa)), interpolate = FALSE)
img2 <-  rasterGrob(as.raster(readJPEG(p_epa)), interpolate = FALSE)

lay = rbind(c(1,2),
            c(1,3))
grid.arrange(img_map, img1, img2,
             layout_matrix = lay,
             widths = c(3,1))
```

```{r animate purpleair adoption, eval=F, include=F}
##Source: https://github.com/thomasp85/gganimate
library(ggplot2)
library(gganimate)
library(gapminder)
## Warning: package 'gapminder' was built under R version 3.4.4
library(gifski)
## Warning: package 'gifski' was built under R version 3.4.4

ggplot(gapminder, aes(gdpPercap, lifeExp, size = pop, colour = country)) +
  geom_point(alpha = 0.7, show.legend = FALSE) +
  scale_colour_manual(values = country_colors) +
  scale_size(range = c(2, 12)) +
  scale_x_log10() +
  facet_wrap(~continent) +
  # Here comes the gganimate specific bits
  labs(title = 'Year: {frame_time}', x = 'GDP per capita', y = 'life expectancy') +
  transition_time(year) +
  ease_aes('linear')

month_counts = pa_data2 %>%
  group_by()

dates <- as.Date(df1$date, '%d/%m/%Y')
yr <- year(dates)
monyr <- as.yearmon(dates)
lst <- lapply(list(dates, yr, monyr), function(x) 
       transform(df1, Count=ave(seq_along(x), x, FUN= length)))
names(lst) <- paste0('newdf', seq_along(lst))
list2env(lst, envir=.GlobalEnv)

```

@@

\begin{itemize}%[itemsep=1.5em]
\item \textbf{Estimate missing pollution observations:} Use PurpleAir data to predict EPA pollution at missing times
\item \textbf{Estimate bias of reported EPA pollution:} difference between predicted pollution at missing times and reported pollution at nonmissing times.
\item \textbf{Estimate counties' counterfactual attainment status:} Include estimated missing pollution data.
\end{itemize}

@@

\[
EPA_{i,t} = \gamma_{i,0} + \sum\limits_{j\in J_i}\sum\limits_{k=1}^7\gamma_{j,k} PA_{j,t} \cdot Winddir_{i,t,k} + u_{i,t}
\]

- Analysis done at the month and quarter level; suppressing that subscript.

- $t$ is a unique hour within a given month or quarter.

- EPA monitor $i$ at time $t$ reads PM2.5 pollution $EPA_{i,t}$.

- For each EPA monitor $i$, there are $J_i$ Purple Air monitors within a 10-mile radius.

- Purple Air monitor $j\in J_i$ at time $t$ reads PM2.5 pollution $PA_{j,t}$.

<!-- [insert diagram of two EPA monitors with PA monitors surrounding them] -->

- $Winddir_{i,t,k}$ is a wind direction indicator; 1 if the prevailing wind near station $i$ at time $t$ is in the $k^{th}$ bucket (of 8 buckets).

<!-- [insert diagram of one EPA monitor and surrounding PA monitors, with wind directions] -->

# Models: Hour-by-Day-of-week Bias of Missing EPA Monitor Pollution Data
\def\mm{\mathcal{M}}
\def\nn{\mathcal{N}}
\vspace{-1em}
\begin{align*}
\intertext{\textbf{Missingness Bias:}}
Bias_{i,h,d} &= \frac{1}{|\nn_{i,h,d}|}\sum\limits_{t\in \nn_{i,h,d}}EPA_{i,t} 
- \frac{1}{|\mm_{h,d}|}\sum\limits_{t\in \mm_{h,d}}\widehat{EPA}_{i,t}\\
\text{where } \mm_{i,h,d}&= \{t: t\text{ is at hour $h$ and day $d$ and $EPA_{i,t}$ is Missing}\};\\
  \nn_{i,h,d}&= \{t: t\text{ is at hour $h$ and day $d$ and $EPA_{i,t}$ is Non-missing}\} \\
\uncover<2->{\intertext{We can also define the \textbf{algorithm bias}     as the Hour-by-Day-of-week prediction error}
\widetilde{Bias}_{i,h,d} &= \frac{1}{|\nn_{i,h,d}|}\sum\limits_{t\in \nn_{i,h,d}}EPA_{i,t} 
- \frac{1}{|\nn_{h,d}|}\sum\limits_{t\in \nn_{h,d}}\widehat{EPA}_{i,t} \\}
\uncover<3->{\intertext{We can also define $Bias_{j,h,d}$ and $\widetilde{Bias}_{j,h,d}$ for PA monitor $j$ (we'll come back to this).}}
\end{align*}

@@

\begin{align*}
Attain^{annual}_c &= 1 \text{ if \textbf{reported} annual average PM2.5 below 15.0 $\mu$g/m$^3$}^*\\[1em]
Attain^{daily}_c &= 1 \text{ if 98$^{th}$ percentile of \textbf{reported} daily average PM2.5 below 35 $\mu$g/m$^3$}^*\\[1em]
\widehat{Attain}^{annual}_c &= 1 \text{ if \textbf{predicted} annual average PM2.5 below threshold}^*\\[1em]
\widehat{Attain}^{daily}_c &= 1 \text{ if 98$^{th}$ percentile of \textbf{predicted} daily average PM2.5 below threshold}^*
\end{align*}
$^*$averaged over 3 years in NAAQS standard.
[fill in equations and thresholds]

@@

- Assumption: nearby PurpleAir monitors that are good predictors for EPA monitors
during non-missing times will also be good predictors during missing times.

  - Specifically, reasons for EPA data missingness are not correlated with missingness
  or measurement error in PurpleAir data

@@

  - Calculate sum of squared missingness bias and sum of squared algorithm bias for EPA sensor $i$ and PA sensors $j\in J_i$.
  \vspace{1em}
  - Ratio$_k$ = sum of squared missingness bias / sum of squared algorithm bias
  \vspace{1em}
  - p-value = $\frac{\text{\# of PA sensors in $i$'s radius with Ratio$_j$ larger than Ratio$_i$}}{\text{\# of PA sensors in $i$'s radius}}$

<!-- But how does this account for poor prediction of the monitor?
Because the denominator is larger if poor prediction in non-missing observations -->

@@

![PurpleAir correction equation for EPA monitor PM2.5 (RH = relative humidity, also measured by PA monitor)](../../docs/purpleair/PA_correction_equation.png){width=50%}

@@

# Appendix C: PurpleAir Takeup
![Valid Purple Air Monitor Locations, Contiguous United States](../../output/summary_stats/graphs/purple_air_sensor_cum_time_california.png){width=60%}
